Reanalysis Step 7: Missingness Pattern Analysis

### 2025-09-01 reanalysis-7-missingness_patterns.qmd
fix ai idiiocy

```{r identify-groups}
# Check if cohort_id exists, if not use a unique identifier
id_var <- if("cohort_id" %in% names(original_data)) "cohort_id" else "e_mrn_deidentified"

# Alternative approach: identify excluded based on missing outcome
# Patients were excluded from imputation if they had missing outcomes
if ("outcome_va_vi_binary" %in% names(original_data)) {
  # Use missing outcome as the exclusion criterion
  original_data$excluded <- is.na(original_data$outcome_va_vi_binary)
  log_message("Using missing outcome_va_vi_binary to identify excluded patients")
} else if ("outcome_va_logMAR" %in% names(original_data)) {
  # Use missing outcome VA
  original_data$excluded <- is.na(original_data$outcome_va_logMAR)
  log_message("Using missing outcome_va_logMAR to identify excluded patients")
} else {
  # Fall back to comparing with imputed dataset IDs
  # Get the IDs from the first imputed dataset (should be same across all)
  included_ids <- unique(imputed_datasets[[1]][[id_var]])

  # Get all IDs from original data
  all_ids <- unique(original_data[[id_var]])

  # Identify excluded patients (those in original but not in imputed)
  excluded_ids <- setdiff(all_ids, included_ids)

  # Create indicator in original data (TRUE = excluded, FALSE = included)
  original_data$excluded <- original_data[[id_var]] %in% excluded_ids
  log_message(paste("Using ID comparison to identify excluded patients.",
                   "Original IDs:", length(all_ids),
                   "Included IDs:", length(included_ids)))
}

@@

# Summary counts
n_total <- nrow(original_data)
n_included <- sum(!original_data$excluded)
n_excluded <- sum(original_data$excluded)
pct_excluded <- round(100 * n_excluded / n_total, 1)

log_message(sprintf("Total patients: %d", n_total))
log_message(sprintf("Included (complete outcome): %d (%.1f%%)", n_included, 100 - pct_excluded))
log_message(sprintf("Excluded (missing outcome): %d (%.1f%%)", n_excluded, pct_excluded))
```

**Sample Size:**
- Total patients: `r n_total`
- Included (complete outcome): `r n_included` (`r round(100 - pct_excluded, 1)`%)
- Excluded (missing outcome): `r n_excluded` (`r pct_excluded`%)

@@

# Print table with standardized mean differences
table1_print <- print(table1, 
                      smd = TRUE, 
                      showAllLevels = TRUE,
                      formatOptions = list(big.mark = ","),
                      printToggle = FALSE)

@@

# Extract SMDs for plotting
smd_data <- ExtractSmd(table1)
smd_data$Variable <- rownames(smd_data)
smd_data <- smd_data[order(abs(smd_data$`1 vs 2`), decreasing = TRUE), ]

# Identify variables with substantial differences (SMD > 0.1)
substantial_diff <- smd_data[abs(smd_data$`1 vs 2`) > 0.1, ]

# Plot 1: SMD plot
p1 <- ggplot(smd_data, aes(x = reorder(Variable, abs(`1 vs 2`)), y = `1 vs 2`)) +
  geom_point(size = 3, color = ifelse(abs(smd_data$`1 vs 2`) > 0.1, "red", "blue")) +
  geom_hline(yintercept = c(-0.1, 0, 0.1), linetype = c(2, 1, 2), alpha = 0.5) +
  coord_flip() +
  labs(title = "Standardized Mean Differences: Excluded vs Included",
       subtitle = "Red points indicate SMD > 0.1 (substantial difference)",
       x = "Variable",
       y = "Standardized Mean Difference") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))

>>>>

```{r identify-groups}
# Use cohort_id if it exists to identify excluded vs included
if ("cohort_id" %in% names(original_data)) {
  # cohort_id has values "excluded" and "submitted"
  # "excluded" = patients with missing outcomes who were excluded from complete case analysis
  # "submitted" = patients with complete outcomes who were included
  original_data$excluded <- original_data$cohort_id == "excluded"
  log_message("Using cohort_id to identify excluded patients")
  log_message(paste("Cohort distribution:", 
                   sum(original_data$cohort_id == "excluded"), "excluded,",
                   sum(original_data$cohort_id == "submitted"), "submitted"))
} else if ("outcome_va_vi_binary" %in% names(original_data)) {
  # Use missing outcome as the exclusion criterion
  original_data$excluded <- is.na(original_data$outcome_va_vi_binary)
  log_message("Using missing outcome_va_vi_binary to identify excluded patients")
} else if ("outcome_va_logMAR" %in% names(original_data)) {
  # Use missing outcome VA
  original_data$excluded <- is.na(original_data$outcome_va_logMAR)
  log_message("Using missing outcome_va_logMAR to identify excluded patients")
} else {
  stop("Cannot identify excluded patients - no cohort_id or outcome variables found")
}

@@

# Summary counts
n_total <- nrow(original_data)
n_included <- sum(!original_data$excluded)  # FALSE = included
n_excluded <- sum(original_data$excluded)   # TRUE = excluded
pct_excluded <- round(100 * n_excluded / n_total, 1)

log_message(sprintf("Total patients (eyes): %d", n_total))
log_message(sprintf("Included (submitted/complete outcome): %d (%.1f%%)", n_included, 100 - pct_excluded))
log_message(sprintf("Excluded (missing outcome): %d (%.1f%%)", n_excluded, pct_excluded))

# Create summary table
summary_df <- data.frame(
  Group = c("Included (submitted)", "Excluded", "Total"),
  N = c(n_included, n_excluded, n_total),
  Percentage = c(round(100 - pct_excluded, 1), pct_excluded, 100)
)
```

```{r summary-table}
#| echo: false
kable(summary_df,
      caption = "Sample Size Summary",
      col.names = c("Group", "N (eyes)", "Percentage (%)"),
      align = c("l", "r", "r"))
```

@@

# Print table with standardized mean differences
# Label the strata clearly
table1_print <- print(table1, 
                      smd = TRUE, 
                      showAllLevels = TRUE,
                      formatOptions = list(big.mark = ","),
                      printToggle = FALSE)

# Update column names for clarity
colnames(table1_print)[colnames(table1_print) == "FALSE"] <- "Included (n)"
colnames(table1_print)[colnames(table1_print) == "TRUE"] <- "Excluded (n)"

@@

# Extract SMDs for plotting
smd_raw <- ExtractSmd(table1)

# Convert to data frame format
# ExtractSmd returns a list where each element is the SMD for a variable
smd_values <- unlist(smd_raw[1:(length(smd_raw)-1)])  # Exclude the Variable element
variable_names <- names(smd_values)

smd_data <- data.frame(
  Variable = variable_names,
  SMD = smd_values,
  stringsAsFactors = FALSE
)

# Remove NA values
smd_data <- smd_data[!is.na(smd_data$SMD), ]

# Order by absolute SMD
smd_data <- smd_data[order(abs(smd_data$SMD), decreasing = TRUE), ]

# Identify variables with substantial differences (SMD > 0.1)
substantial_diff <- smd_data[abs(smd_data$SMD) > 0.1, ]

# Plot 1: SMD plot
p1 <- ggplot(smd_data, aes(x = reorder(Variable, abs(SMD)), y = SMD)) +
  geom_point(size = 3, color = ifelse(abs(smd_data$SMD) > 0.1, "red", "blue")) +
  geom_hline(yintercept = c(-0.1, 0, 0.1), linetype = c(2, 1, 2), alpha = 0.5) +
  coord_flip() +
  labs(title = "Standardized Mean Differences: Excluded vs Included",
       subtitle = "Red points indicate SMD > 0.1 (substantial difference)",
       x = "Variable",
       y = "Standardized Mean Difference") +
  theme_minimal() +
  theme(plot.title = element_text(size = 14, face = "bold"))
